---
title: "SAKPlot Cheat Sheet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SAKPlot Cheat Sheet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SAKPlot)
library(datasets)
data(iris)
```


This vignette shows some examples of plots that can be generated by the `SAKPlot` package. Each plot is preceded by the code used to generate it, which can be used to learn how to create the plot. 

The plotting order of the groups can be controlled by specifying the `groups` parameter. Not all groups in the data need be included in the analysis or plot.

The colours of all components can controlled using the appropriate parameters. A single colour may be specified, or different colours can be set for each group. 

```{r fig.height=5, fig.width=9}
par(mfrow = c(1, 2), mar = c(5, 4, 3, 6) + 0.1)
d <- SAKDifference(iris, groups = c("versicolor", "setosa"), data.col = "Sepal.Length", group.col = "Species")
SAKPlot(d, main = "Two groups")
par(mar = c(5, 4, 3, 2) + 0.1)
d <- SAKDifference(iris, data.col = "Sepal.Length", group.col = "Species")
SAKPlot(d, violin = "grey50", violin.fill = "grey90", violin.shape = "left", points = SAKTransparent("black", 0.9), main = "Three groups")
```

Error bars on each group may show the 95% confidence intervals of the mean, standard deviation or standard error, and may be plotted regardless of what other components are shown.
 
```{r fig.height=6, fig.width=9}
par(mfrow = c(1, 2))
d <- SAKDifference(iris, data.col = "Sepal.Length", group.col = "Species")
SAKPlot(d, bar = TRUE, error.bars.type = "SD", points = FALSE, violin = FALSE, main = "Bar chart with std. deviation")
SAKPlot(d, error.bars.type = "CI", central.tendency = c("red", "blue", "pink"), points = FALSE, violin = FALSE, main = "Mean and 95% CI")
```

By default, effect sizes are "unstandardised", which means they are in the same units as the data. Alternatively, \code{SAKDifference} can estimate standardised effect sizes, either Cohen's d or Hedges' g, which allows comparisons of effect sizes across data in different units.

```{r fig.height=6, fig.width=9}
par(mfrow = c(1, 2))
d <- SAKDifference(iris, data.col = "Sepal.Length", group.col = "Species", effect.type = "cohens")
SAKPlot(d, main = "Cohen's d")
d <- SAKDifference(iris, data.col = "Sepal.Length", group.col = "Species", effect.type = "hedges", ci.conf = 0.9)
SAKPlot(d, ef.size.violin = FALSE, central.tendency = "blue", violin.trun = FALSE, main = "Hedges' g, 90% CI")
```

```{r fig.height=5, fig.width=7}
par(mar = c(5, 4, 3, 5) + 0.1)
# Add in some fake sex data to the insulin data set
data <- cbind(insulin, Sex = sample(c("Male", "Female"), nrow(insulin), replace = TRUE))
# Thin it the data so individual symbols are visible
data <- data[data$id %in% 1:15, ]

d <- SAKDifference(data, "sugar", "treatment", "id", groups = c("Before insulin" = "before", "After insulin" = "after"), na.rm = TRUE)
SAKPlot(d, 
        left.ylab = "Blood sugar level",
        violin.shape = c("left", "right"), violin.dx = c(-0.055, 0.055), violin.width = 0.3, 
        points = "black",
        points.params = list(bg = ifelse(data$Sex == "Female", "red", "blue"), 
                             pch = ifelse(data$Sex == "Female", 21, 24)), 
        ef.size.pch = 19,
        ef.size.violin = "blue",
        ef.size.violin.shape = "full",
        central.tendency = FALSE,
        main = "Customised plot")
```

```{r fig.height=4, fig.width=7}
par(mar = c(5, 4, 3, 1) + 0.1)
set.seed(1)
data <- data.frame(Measurement = unlist(lapply(c(10, 12, 8, 8.5, 9, 9.1), function(m) rnorm(20, m))),
                   Group = rep(paste0("g", 1:6), each = 20),
                   Id = rep(1:6, 20))
d <- SAKDifference(data, 1, 2, 3, groups = c("Control 1" = "g1", "Treatment 1" = "g2", "Control 2" = "g3", "Treatment 2" = "g4", "Control 3" = "g5", "Treatment 3" = "g6"), 
                   contrasts = "g2 - g1, g4 - g3, g6 - g5")
SAKPlot(d, box = TRUE, group.dx = c(0.2, -0.2), paired = FALSE, points = FALSE, box.pars = list(boxwex = 0.4), ef.size.violin = FALSE, main = "Group offsets")
```

```{r fig.height=5, fig.width=6}
par(mar = c(5, 4, 3, 9) + 0.1)
effectSizes <- c("No effect" = 0, "Large positive effect" = 0.8, "Huge positive effect" = 2)
d <- SAKDifference(damselfly, 1, 3, effect.type = "cohen", groups = c("Juvenile" = "juvenile", "Adult" = "adult"), na.rm = TRUE)
SAKPlot(d, violin.shape = c("left", "right"), violin.dx = c(-0.05, 0.05), violin.width = 0.3, main = "Effect size interpretation, right", ef.size.ticks = effectSizes, ef.size.label = "", ef.size.las = 1)

par(mar = c(5, 9, 3, 1) + 0.1)
SAKPlot(d, box = TRUE, ef.size.ticks = effectSizes, ef.size.position = "below", ef.size.las = 1, main = "Effect size interpretation, below", ef.size.label = "")
```


