---
title: "Estimation plotting with `SAKPlot`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation plotting with `SAKPlot`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> “The primary product of a research inquiry is one or more measures of effect size, not _p_ values” (Cohen, 1990, p. 1310)

```{r setup}
library(SAKPlot)
```

## What is `SAKPlot`?

`SAKPlot` is an R package that aims to simplify the estimation and communication of one type of effect size: differences in means. It provides a function, `SAKDifference`, to estimate raw or standardised mean differences. It provides `print` methods to display the calculated differences. Perhaps most importantly, it provides a single function, `SAKPlot`, that will create a plot for visualising both the original data and the effect size, providing a diverse range of display options.



### Estimating mean differences

#### Multiple groups, multiple comparisons

The `contrasts` parameters to `SAKDifference` and `SAKPlot` ...

### Visualising mean differences

#### Comparison of different types of error bars

```{r fig.height=4, fig.width=7}
par(mar = c(2.5, 4, 2, 0.5) + 0.1)
set.seed(3)
df <- data.frame(var = c(rnorm(5, 50, 17), rnorm(20, 50, 17), rnorm(80, 50, 17)),
                 group = c(rep("N = 5", 5), rep("N = 20", 20), rep("N = 80", 80)))
d <- SAKDifference(df, 1, 2, groups = c("N = 5", "N = 20", "N = 80"))
SAKPlot(d, error.bars.type = "CI", error.bars.cross.width = 0.03, violin = FALSE, points.method = "jitter", central.tendency.dx = .2, ef.size = FALSE, central.tendency.params = list(cex = 1), main = "Error bar types: CI, SE, SD", left.ylab = "Dependent variable")
abline(h = 50, col = "#00000020")
SAKPlot(d, add = TRUE, error.bars.type = "SE", error.bars.cross.width = 0.03, violin = FALSE, points = FALSE, central.tendency.dx = .3, ef.size = FALSE, central.tendency.params = list(cex = 1))
SAKPlot(d, add = TRUE, error.bars.type = "SD", error.bars.cross.width = 0.03, violin = FALSE, points = FALSE, central.tendency.dx = .4, ef.size = FALSE, central.tendency.params = list(cex = 1))

```


## References

Cohen, J. (1990). Things I have learned (so far). American psychologist, 45(12), 1304. 
