---
title: "Estimation plotting with `SAKPlot`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation plotting with SAKPlot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> “The primary product of a research inquiry is one or more measures of effect size, not _p_ values” (Cohen, 1990, p. 1310)

```{r setup}
library(SAKPlot)
```

## What is `SAKPlot`?

`SAKPlot` is an R package that aims to simplify the estimation and communication of one type of effect size: differences in means. It provides a function, `SAKDifference`, to estimate raw or standardised mean differences. It provides `print` methods to display the calculated differences. Perhaps most importantly, it provides a single function, `SAKPlot`, that will create a plot for visualising both the original data and the effect size, providing a diverse range of display options.



### Estimating mean differences



### Visualising mean differrences

## References

Cohen, J. (1990). Things I have learned (so far). American psychologist, 45(12), 1304. 
